Artificial ecosystem selection to deduce the mapping between microbial community structure and ecosystem function

### 2021-07-01 Manuscript/manuscript.Rmd
revise introduction

## Determining the mapping between microbial community structure and ecosystem function

Microbial communities mediate ecosystem nutrient cycling via decomposition,
greenhouse gas cycling, carbon storage, and other processes. Despite the
central role of bacteria in performing these processes, we have not found a
consistent relationship between the composition of a microbial community and
the rate of an ecosystem process across habitats. Typically, research in this
area has followed the model of macroorganismal biodiversity-ecosystem function
studies such as grassland productivity studies that measure or manipulate the
diversity of plant species and correlate that with the rate of primary
productivity. By analogy, microbial ecologists use microbial sequencing data to
create taxonomic relative abundance tables and then infer diversity or relative
abundance from the whole archaeal/bacterial community and correlate that with
the rate of a microbial process. Alternatively, they estimate the abundance of
a putative functional gene (such as *pmoA* for methane oxidation) and correlate
abundance with function. These approaches only rarely show significant,
positive relationships between biodiversity and ecosystem function. In my first
chapter published in Phil Trans I lay out philosophically why this approach
shouldn’t work – namely that marker gene sequences are not analogous to plant
species counts, but are more analogous to measurements of gene markers as in
evolutionary biology studies.

In evolutionary biology, genotype-phenotype mapping provides a way to
understand the fitness effects of a genotype. While selection typically acts on
phenotype, the genotypic changes are the fundamental driver of phenotypic
differences and thus fitness differences (Stadler and Stephes 2005). A similar
analogy can be drawn for biodiversity-ecosystem function relationships. Traits
of a community are what drive community assembly, not the taxa or genotypes
themselves (Inkpen et al. 2017). However, as ecologists, we wish to understand
the effects of species on ecosystem function. Therefore, in a similar way that
evolutionary biologists want to deduce the genotype-phenoype map to understand
the genetic basis of fitness, community ecologists must describe the "mapping"
between community genotype and ecoystem function to understand what drives
communty assembly, and ultimately, ecosystem function.

The relationship that ecologists have described between plant diversity and
productivity represents a relatively simple map between taxonomy and function.
The number of unique members of a functional group (i.e., plant species) is
correlated with the rate of the function performed by that group. However, for
microbial communities that mapping is not quite as clear. For many microbial
traits, there is not a clear coupling between taxonomy and function, except for
in the case of highly conserved traits involving many genes (Martiny et al.
2013). In addition, most microbial taxa have not been sampled and so we don't
have a full picture of which taxa are members of a particular functional group
(citation). Put another way, it's much harder to identify methanotrophic
prokaryotes in soil than it is to identify plants in a prairie. Finally, even
for highly conserved traits with a putative functional marker, for example
methane oxidation and the gene *pmoA*, functional group abundance is not
correlated with the rate of the function performed by that group in most
ecosystems (Rocca et al. 2016). Because of these differences between
macroorganisms and microorganisms, it may be more useful to approach microbial
biodiversity-ecosystem function in a way analogus to genotype-phenotype mapping
in organisms (Morris et al. 2020).

One approach to deducing the genotype-phenotype map in organisms is through
artificial selection (Fuller et al. 2005). In this approach, selection for a
trait is imposed in a contrived laboratory environment and compared to
laboratory environment without selection on that trait. By comparing the
genetic respones between these two treatments, you can remove the
environment-fitness connection that would be present in a comparative study and
identify genes that are evolving in response to selection only on the trait of
interest. It has been shown that whole communities can respond to artifical
selection on traits of the ecosystem (Goodnight et al. 1997, Williams and
Lenton 2007). And it has been proposed that this could be a suitable approach
for identify communities of organisms that are capable of performing a
particular function at a high rate (Swenson et al. 2000). Therefore, artificial
ecosystem selection is a potentially fruitful approach to identifying the map
between biodiversity and ecosystem function in microbial communities.

To identify the map between microbial community composition and ecosystem
function, we imposed artificial ecosystem selection on methane oxidation rate
of whole soil ecosystems. We prepared two lines of replicate ecosystems: one in
which we imposed positive selection on methane oxidation by only using the top
performing ecosystems to generate the next set of ecosystems and one in which
we imposed neutral selection where a random set of ecosystems were chosen to
generate the next set of ecosystems. We repated this process over multiple
passages and then compared the resultant metacommunities in terms of their rate
of ecosystem function and the composition of their communities.

With this experiment, we asked several questions. First, is there a response to
selection on methane oxidation by soil ecosystems? And how strong is the
response to selection? (i.e., how "heritable" is methane oxidation as a
community trait in terms of the phenotypic resemblence of ecosystems to the
ecosystems that generated them?) If methane oxidation is heritable, then what
is the nature of the map between microbial community composition and methane
oxidation? Is it the diversity of methane oxidizers, as it is for plants and
productivity? Or is there a more complex map involving upstream metabolic
regulation?

Things this Intro doesn't yet discuss:

1. What a more complex map might look like. For example, maybe pmoA and methanotrophy aren't correlated, but maybe understand methanogenesis or nitrogen cycling is helpful.

2. The idea of "heritability" for communities/ecosystems and why that relates to environmental vs. genetic control over ecosystem function.

>>>>

One of the fundamental questions in Ecology is To what extent does biodiversity
drive variation in ecosystem functioning? Variation in abiotic conditions
regulates the rate of ecosystem processes, but it remains unclear to what
extent variation in community structure regulates the rate of ecosystem
processes. Many ecosystem functions are mediated by organisms. In particular,
microorganisms such as bacteria mediate the major biogeochemical cycles such
as the carbon and nitrogen cycles. Despite the central role of bacteria in
performing these processes, we have not found a consistent relationship between
microbial community structure and the rate of any ecosystem process. In order
to understand the importance of biodiversity to ecosystem functioning and to
better predict and manage specific ecosystem processes, we need to determine
the relationship between microbial community structure and ecosystem function.

In macroorganismal communities, there appears to be a relationship between the
diversity of a community and the resilience of ecosystem functioning to
environmental perturbations (Tilman, ????... Hooper). Typically, this is the
diversity (or more specifically, the richness) of a particular functional
group, such as photosynthetic plants. Similar studies in microbial communities
rarely find a relationship between diversity and function (Graham, ). However,
these studies typically define diversity much more broadly than plant diversity
studies by looking at the phylogenetic diversity of the entire bacterial and
archaeal community, not just of a particular functional group. Studies that
focus more narrowly on a particular ecosystem process and a functional group
involved in that process still rarely find a significant correlation between
the abundance of a functional group (or a functional protein-coding gene) and
the rate of the corresponding process (Rocca, ????). For certain ecosystem
functions in certain ecosystems, these relationships can be informative. For
example, the abundance of a single methanogenic species (*G. species*) is
correlated with methane emissions in a permafrost ecosystem (cite). In addition,
bacterial and archaeal community composition is correlated with methane
emissions in forests and pastures of the Brazilian Amazon (Meyer et al. 2020).
Therefore, there appears to be some indication that microbial community
structure drives variation in methane cycling in soils.

To further investigate the relationship between microbial community structure
and the rate of methane cycling in soils, we want to determine which organisms
are important for regulating methane cycling. This ecological question is
analogous to the problem of genotype-phenotype mapping in quantitative genetics
(Morris 2020). Variation in an organimsal trait is partially determined by
environmental conditions and partially determined by genetic varition. One goal
of quantitative genetics is to determine the proportion of the variation
attributable to genetics and to identify which alleles regulate the phenotypic
trait. These goals are commonly achieved through artificial selection
experiments and association mapping studies, such as genome-wide association
studies. To determine the mapping between microbial community structure and
ecosystem function, we applied these approaches of genotype-phenotype mapping
to whole microbial communities. In this case, the "alleles" are different
microbial taxa and the "phenotype" is the rate of ecosystem function at the
whole-ecosystem level.

To do this, we performed an artificial ecosystem selection experiment to
identify communities of bacteria and archaea that collectively perform a high
rate of methane oxidation (Swenson et al. 2000).  (Goodnight et al. 1997,
Williams and Lenton 2007)we generated twenty-four soil ecosystems in the lab
using a sterilized potting mix as the substrate. We then inoculated these
ecosystems with a small amount of living soil to generate variation in
community structure across the twenty-four ecosystems. These ecosystems were
then assigned to one of two selection treatments: neutral selection with
ecosystems chosen for reproduction at random or positive selection with
ecosystems chosen for reproduction based on their methane oxidation rate. Each
ecosystem was maintained at 1000 ppm CH4 over several weeks to allow for
colonization and growth of the microbial community and to enrich for methane
oxidizers. We then determined the methane oxidation rate for each ecosystem and
selected three jars in each treatment to inoculate the next set of twenty-four
jars. This process continued for five passages. At the end of the experiment,
we extracted DNA from soils in the second and fifth passage for both the
neutral and positive treatments. By comparing the response to selection within
the community between the two treatments, we can identify microbial species
that contribute to variation in ecosystem methane emissions.

This approach is powerful because by using a common soil substrate and
maintaining a constant headspace concentration of methane, we can eliminate
much of the environmental variation that would be present in an observational
study in the wild. In addition, by applying selection on methane oxidation rate
in the positive treatment, we can enrich for taxa involved in methane cycling
making it easier to perform association mapping between microbial taxa and
ecosystem methane emissions. In this paper, we address whether there is a
response to selection on methane oxidation rate at the whole ecosystem level. We
then quantify the amount of variation in methane oxidation attributable to
variation in community composition within our laboratory environment. Finally,
we identify microbial markers of soil methane cycling using association mapping
in order to deduce the mapping between microbial community structure and
ecosystem function.
